Estimate of the Neural Network Dimension Using Algebraic Topology and Lie Theory\thanks{The code can be found at: \href{https://github.com/karhunenloeve/ neural-topology-lie}{https://github.com/karhunenloeve/neural-topology-lie}.}

### 2020-07-14 llncs2e/paper12.tex
final

A metric describes the geometric properties of a manifold. The change of coordinate systems and the learned metric tensor of a smooth manifold during backpropagation was formalized \cite{HauserR17}, which shows, that we operate often on a model of a Lie Group. Further, it has been shown that dense neural networks cannot approximate all functions arbitrarily precisely \cite{Johnson19}. Thereupon neural networks with residual connections were investigated. Residual nets add the output of one layer to a deeper one, bypassing some of the layers in between. Indeed, they are universal function approximators \cite{LinJ18}. Inspired by finite difference methods, such a residual layer was defined as a forward or backward difference operator on a partition of layers. The state space dimension with residual connections is homeomorphic to $\mathbb{R}^{k \cdot n}$, where $n = \text{dim} \; M$ and $k$ is the number of times the difference operator was used \cite{HauserGJR19}. Such a layer is able to embed in $k\cdot n$ dimensional Euclidean space. The smooth manifold perspective inspired us to investigate its invariants leading into the theory discussed in this paper.

\subsubsection*{Embedding dimension of neural networks} Cybenko showed that dense neural networks with one hidden layer can approximate any continuous function with a bounded domain with arbitrarily small error \cite{Cybenko92}. The upper bound of the embedding dimension for ReLU nets was recently determined to be $(n+4)$, where $n$ is the sample dimension \cite{RaghuPKGS17}. Bartlett et al. show an almost tight bound for VC dimensions. A neural network with $W$ weights and $L$ layers is bounded by $O(WL \log W)$ VC dimensions but has $\Omega(WL \log W/L)$ VC dimensions \cite{BartlettHLM19}. Its depth was determined as a parameter depending on the moduli of continuity of the function to be approximated \cite{LinJ18}.

@@

\subsection{Smooth manifolds}
A smooth structure is a stronger condition than a topological manifold which can be described by a family of continuous functions. This assumption is justified by the theorem of Stone-Weierstrass which proves that every continuous function can be approximated arbitrarily exactly by a smooth one \cite{stone1948generalized}. Smooth manifolds are described by a family of local coordinate maps $\varphi: U \rightarrow \varphi(U) \subseteq \mathbb{R}^n$ which are homeomorphic to a subset of $\mathbb{R}^{n}$ and cover an open neighborhood $(U_i,\varphi_i)$ in $M$. A family of charts, which covers $M$, is called \emph{atlas} on $M$, i.e. $\mathscr{A} = (U_i,\varphi_i)_{i \in I}$ \cite{lee2013smooth}. The functions $x_1, \cdots, x_n: U \rightarrow \mathbb{R}$ are called local coordinates $\varphi(p) = (x_1(p), \cdots, x_n(p))$. If the atlas is maximal in terms of inclusion, then it is a \emph{differentiable structure}. Each atlas for $M$ is included in a maximal atlas. Due to the differentiable structure, the maps in $\mathscr{A}$ are also compatible with all maps of $\mathscr{A}_{\text{max}}$. It is sufficient to describe smooth manifolds by a non maximal atlas.

@@

\subsubsection{Lie groups} The study of group theory deals with symmetries which can be expressed algebraically. A pair $(M,\bullet)$, consisting of a map $\bullet: M\times M \rightarrow M$ is called group if the map is associative, i.e. $x \bullet (y \bullet z) = (x \bullet y) \bullet z$ and has a neutral element so that $x \bullet e_{M} = x$. We also require an inverse element to each element of the group, i.e. $x \bullet x^{-1} = e_{M}$. A Lie group is a smooth manifold which is equipped with a group structure such that the maps $\bullet:M \times  M \rightarrow  M, (x,y) \mapsto xy, $ and $\imath: M \rightarrow  M, x \mapsto x^{-1}$ are smooth. We call a space \emph{connected} if it cannot be divided into disjoint open neighbourhoods. A group is called Abelian or commutative when all elements under the group operation commute.

We apply a theorem from Lie theory, which states that each connected Abelian Lie group $(M,\bullet)$ of dimension $\dim M = n$ is isomorphic to a product space $\mathbb{R}^p \times \mathbb{T}^q$ with $p+q =n$, for a proof we refer to \cite[p.~116]{onishchik1993lie}. The $q$-torus $\mathbb{T}^q$ is a surface of revolution. It moves a curve around an axis of rotation. These axes are given by $1$-spheres, such that the $q$-torus is a product space $\mathbb{T}^q = \mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q$. The initial decomposition of connected commutative Lie groups can be further simplified into $M \simeq \mathbb{R}^p \times (\mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q)$. Recall, that the $(n-1)$-sphere is given by $\mathcal{S}^{n-1} := \{x \in \mathbb{R}^n \; \vert \; \norm{x}_2 = 1\}$. Thus, $\mathcal{S}^1$ can be embedded into $\mathbb{R}^2$. Next, we derive how many dimensions are at least needed for a suitable embedding. For each $1$-sphere we count two dimensions and for each real line accordingly one. Finally, we have to estimate from data how the decomposition may look like to yield the topology of the data manifold.

@@

\input{figure1.tex}
\subsubsection{Simplicial complexes} A \emph{simplicial complex} $K$ is a set of simplices, such that any face of a simplex of $K$ is a simplex of $K$ and the intersection of any two simplices of $K$ is either empty or a common face of both \cite[p.~11]{boissonnat2018geometric}. We require every $(k+1)$-element, with $k < n$ to be a $k$-simplex of $[\sigma] \subset K$. This defines the simplicial complex as a topological space. These properties ensure that simplices are added to the complex in a way, such that they are \emph{glued} edge to edge, point to point, and surface to surface. Thus, every $n$-simplex has $n+1$ distinct vertices and no other $n$-simplex has the same set of vertices. This is a unique combinatorial description of vertices together with a collection of sets $\{[\sigma]_1, \cdots, [\sigma]_m\}$ of $k$-simplices, which are $k+1$ element subsets of $K$ \cite[p.~107]{hatcher2002algebraic}. This set system can be realized geometrically with an injective map into Euclidean space.

@@

The $k$-th chain groups contain two more Abelian subgroups that behave normal to it. First, the so-called cycle groups $\langle Z_k, +\rangle$, which are defined as $Z_k := \ker \partial_k = \{[\sigma] \in C_k \; \vert \; \partial_k [\sigma] = \emptyset \}$. This follows the intuition of all elements that form a loop, i.e. have the same start and end point. Some of these loops have the peculiarity of being a boundary of a subcomplex. These elements form the boundary group $\langle B_k, +\rangle$, defined by $B_k := \text{im } \partial_{k+1} = \{[\sigma] \in C_k \; \vert \; \exists [\tau] \in C_{k+1}: [\sigma] = \partial_{k+1}[\tau]\}$. Now we define the homology groups as quotient of groups. Homology -- analogous to homotopy theory -- gives information about connected components, loops and higher dimensional holes in the simplicial complex:
\begin{equation}
    H_k(K) := \frac{\ker \partial_k C_k(K)}{\text{im } \partial_{k+1} C_{k+1}(K)} = \frac{Z_k(K)}{B_k(K)}.
\end{equation}

\subsubsection{Homological persistence} Examining the homology groups of a set of points gives little information about the structure of a dataset. Instead, we are interested in a parametrization of the simplicial complex as geometric realization in which the homology groups appear and disappear again. For this purpose we consider all possible subcomplexes that form a \emph{filtration} over the point set $X$. We denote $K^{\epsilon} := K^{\epsilon}(X)$. Depending on how we vary the parameter $\epsilon_i$ of the chosen simplicial complex, the following sequence is generated, connected by inclusion and starting with the empty set cf. Fig. \ref{figure1} b):
\begin{align}
    \emptyset &= K^{\epsilon_0} \subset K^{\epsilon_1} \subset \cdots \subset K^{\epsilon_{n+1}}= K,\\
    K^{\epsilon_{i+1}} &= K^{\epsilon_i} \cup [\sigma]^{\epsilon_{i+1}}, \quad \text{for} \; i \in \{0, \cdots, n-1\}.
\end{align}
The filtration has a discrete realization with a fixed $\epsilon = \epsilon_{i+1}-\epsilon_{i} = \min(x,y)$ for all $x,y \in X$. Through filtration we are able to investigate the homology groups during each step of the parameterization. We record when elements from a homology group appear and when they disappear again. Intuitively speaking, we can see when $k$-dimensional holes appear and disappear in the filtration. We call this process \emph{birth} and \emph{death} of topological features. Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}, see Fig. \ref{figure1} c)3. The Betti numbers $\beta_k$ are defined by $\text{rank } H_k$. We write $H^{\epsilon_i}_k$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\epsilon_i$. Then $H^{\epsilon_{i}}_{k} \rightarrow H^{\epsilon_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, for a proof we refer to \cite{edelsbrunner2008persistent}:
\begin{align}
0 = H^{\epsilon_{0}}_k \rightarrow H^{\epsilon_{1}}_k \rightarrow \cdots \rightarrow H^{\epsilon_{n}}_k \rightarrow H^{\epsilon_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\epsilon_i}$ or appear before and die after spanning $K^{\epsilon_{i+1}}$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\epsilon_{i},\epsilon_{j}}_k = (\beta_k^{\epsilon_{i},\epsilon_{j-1}} - \beta_k^{\epsilon_i,\epsilon_{j}})-(\beta_k^{\epsilon_{i-1},\epsilon_{j-1}}-\beta_k^{\epsilon_{i-1},\epsilon_{j}}),
\end{equation}
for the $k$-th homology group and index pairs $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \infty$ with indices $i \leq j$. The Euclidean space is extended, as the very first connected component on the filtration remains connected. Thus, we assign to it infinite persistence, corresponding to the second coordinate $\epsilon_{j+1}$. The first term counts elements born in $K^{\epsilon_{j-1}}$ and which vanish entering $K^{\epsilon_{j}}$, while the second term counts the representatives of homology classes before $K^{\epsilon_{j}}$ and which vanish at $K^{\epsilon_{j}}$. The $k$-th persistence diagram is then defined as
\begin{equation}
\text{PH}_{k}(X) := \left\{(\epsilon_i, \epsilon_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\epsilon_{i},\epsilon_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}

\subsubsection{Persistence landscapes} Persistence landscapes give a statistical summary of the topology of a set of points embedded in a given topological manifold \cite{bubenik2015statistical}. Looking at the points $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2}$ on the $k$-th persistence diagram $\text{PH}_{k}(X)$, one associates a piecewise linear function $\lambda^{\epsilon_i}_{\epsilon_{j+1}}: \mathbb{R} \rightarrow [0,\infty)$ with those points:
\begin{align}
    &\text{If } x \not\in (\epsilon_i,\epsilon_{j+1}), \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = 0,\\
    &\text{if } x \in \left(\epsilon_{i},(\epsilon_i + \epsilon_{j+1})/2 \right], \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = x - \epsilon_{i} \; \text{and}\\
    &\text{if } x \in \left((\epsilon_i + \epsilon_{j+1})/2,\epsilon_{j+1} \right), \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = \epsilon_{j+1} - x.
\end{align}
The summaries for the general persistence diagram are the disjoint union of the $k$-th persistence diagrams $\text{PH}(X) := \coprod_{i=0}^{k} \text{PH}_i(X)$. A persistence landscape $\text{PL}(X)$, contains the birth-death pairs $(\epsilon_i,\epsilon_{j+1})$, for an $i,j \in \{1, \cdots, n\}$ and is the function sequence $\Lambda_k : \mathbb{R} \rightarrow [0,\infty)$ for a $k \in \mathbb{N}$, where $\Lambda_k(x)$ denotes the $k$-th greatest value of $\lambda^{\epsilon_i}_{\epsilon_{j+1}}(x)$, see Fig. \ref{figure1} c)4. Thus, $\Lambda_k(x) = 0$ if $k > n$. $\text{PL}(X)$ lies in a completely normed vector space, suited for statistical computations.

@@

In practice, neural networks are used with a different number of neurons per layer. However, a layer can represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map does not change. Thus, the dimension of the data manifold is equal to the width of the smallest possible layer of a neural network \cite{HauserR17}.

@@

We count the representatives of homology classes in the persistence landscapes to derive the dimension of the sought manifold. We get for $\mathcal{S}^1$:
\begin{align}
    H_0(\mathcal{S}^1) &= H_1(\mathcal{S}^1) = \mathbb{Z},\\
    H_i(\mathcal{S}^1) &= 0, \; \text{for all } i \geq 2.
\end{align}
Note, in Eq. \ref{homologykunneth} terms remain only for indices $i_{j} \in \{0,1\}$. Thus, we get
\begin{align}
    H_0(\mathbb{R}^p) &\simeq\mathbb{Z}^{p},\\
    \label{determineq}
    H_k(\mathbb{T}^{q}) &\simeq H_k(\mathcal{S}^{1}_{1} \times \cdots \times \mathcal{S}^{1}_{q}) \simeq \mathbb{Z}^{{q\choose{k}}},
\end{align}
where $p$ indicates the number of connected components. The number of local maxima for a homology group in the persistence landscape is the cardinality of the set of points $(\epsilon_i,\epsilon_{j+1})$ with derivatives $d\lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = 0 \; \text{and} \; d^2 \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) < 0$, see Fig. \ref{figure2} b) c). Also, it is the solution for the binomial coefficient:
\begin{align}
     {q \choose k} = \frac{q}{1} \cdot \frac{q-1}{2} \cdots \frac{q-(k-1)}{k} = \prod_{i=1}^{k} \frac{q+1-i}{i}.
\end{align} 
We count in $PL(X)$ the elements of the $0$th homology group. The amount of elements from higher homology groups correspond to $q$ (Eq. \ref{determineq}) and are computed using Monte-Carlo methods for an (approximate) integer solution, see Tab. \ref{figure3}.

@@

Applying Künneth's theorem, the homology groups of the topological factor spaces could be connected with the ones of their product space. Using persistence landscapes the elements originating from some homology groups on the filtration were estimated. With numerical experiments we predicted near ideal embedding dimensions and could confirm, see Fig. \ref{figure2}, that a neural embedding above the treshold delivers a loss function on the validation data set with small fluctuations and best reconstruction results. We pose following open research questions:
\begin{itemize}
    \item The neural layers do not have the explicit structure of an Abelian Lie group. How can spherical CNNs \cite{CohenGKW18} be used to always represent a product space $\mathbb{R}^p \times \mathbb{T}^q$ from layer to layer in order to operate on the proposed Lie group?
    \item This approach can be applied to any kind of manifold, as far as its minimal representation is known. Using decompositions, one may generalize this result. What decomposition allows to neglect the connectedness?
    \item What do other representations of manifolds with a more complicated structure look like? For example, can manifolds like the Klein bottle be learned that are not orientable? And if so, how can one decide in advance whether or not the approximation to an orientable manifolds makes more sense than to a non orientable one?
\end{itemize}

>>>>

A metric describes the geometric properties of a manifold. The change of coordinate systems and the learned metric tensor of a smooth manifold during backpropagation was formalized \cite{HauserR17}, which shows, that we operate often on a model of a Lie Group. Further, it has been shown that dense neural networks cannot approximate all functions arbitrarily precisely \cite{Johnson19}. Thereupon, neural networks with residual connections were investigated. Residual networks add the output of one layer to a deeper one, bypassing some of the layers in between. Indeed, they are universal function approximators \cite{LinJ18}. Inspired by finite difference methods, such a residual layer was defined as a forward or backward difference operator on a partition of layers. The state space dimension with residual connections is homeomorphic to $\mathbb{R}^{k \cdot n}$, where $n = \text{dim} \; M$ and $k$ is the number of times the difference operator was used \cite{HauserGJR19}. Such a layer is able to embed in $k\cdot n$ dimensional Euclidean space. The smooth manifold perspective inspired us to investigate its invariants leading into the theory discussed in this paper.

\subsubsection*{Embedding dimension of neural networks} Cybenko showed that dense neural networks with one hidden layer can approximate any continuous function with a bounded domain with arbitrarily small error \cite{Cybenko92}. The upper bound of the embedding dimension for \texttt{ReLU} networks was recently determined to be $(n+4)$, where $n$ is the sample dimension \cite{RaghuPKGS17}. Bartlett et al. show an almost tight bound for VC dimensions. A neural network with $W$ weights and $L$ layers is bounded by $O(WL \log W)$ VC dimensions but has $\Omega(WL \log W/L)$ VC dimensions \cite{BartlettHLM19}. Its depth was determined as a parameter depending on the moduli of continuity of the function to be approximated \cite{LinJ18}.

@@

\subsection{Smooth manifolds}
A smooth structure is a stronger condition than a topological manifold which can be described by a family of continuous functions. This assumption is justified by the theorem of Stone-Weierstrass which proves that every continuous function can be approximated arbitrarily exactly by a smooth one \cite{stone1948generalized}. Smooth manifolds are described by a family of local coordinate maps $\varphi: U \rightarrow \varphi(U) \subseteq \mathbb{R}^n$ which are homeomorphic to a subset of $\mathbb{R}^{n}$ and cover an open neighborhood $(U_i,\varphi_i)$ in $M$. A family of charts, which covers $M$, is called \emph{atlas} on $M$ \cite{lee2013smooth}, i.e. $\mathscr{A} = (U_i,\varphi_i)_{i \in I}$. The functions $x_1, \cdots, x_n: U \rightarrow \mathbb{R}$ are called local coordinates $\varphi(p) = (x_1(p), \cdots, x_n(p))$. If the atlas is maximal in terms of inclusion, then it is a \emph{differentiable structure}. Each atlas for $M$ is included in a maximal atlas. Due to the differentiable structure, the maps in $\mathscr{A}$ are also compatible with all maps of $\mathscr{A}_{\text{max}}$. It is sufficient to use a non maximal atlas.

@@

\subsubsection{Lie groups} The study of group theory deals with symmetries which can be expressed algebraically. A pair $(M,\circ)$, consisting of a map $\circ: M\times M \rightarrow M$ is called group if the map is associative, i.e. $x \circ (y \circ z) = (x \circ y) \circ z$ and has a neutral element so that $x \circ e_{M} = x$. We also require an inverse element to each element of the group, i.e. $x \circ x^{-1} = e_{M}$. A Lie group is a smooth manifold which is equipped with a group structure such that the maps $\circ: M \times  M \rightarrow  M, (x,y) \mapsto xy, $ and $\imath: M \rightarrow  M, x \mapsto x^{-1}$ are smooth. We call a space \emph{connected} if it cannot be divided into disjoint open neighbourhoods. A group is called Abelian or commutative when all elements under the group operation commute.

We apply a theorem from Lie theory, which states that each connected Abelian Lie group $(M,\circ)$ of dimension $\dim M = n$ is isomorphic to a product space $\mathbb{R}^p \times \mathbb{T}^q$ with $p+q =n$, for a proof we refer to \cite[p.~116]{onishchik1993lie}. The $q$-torus $\mathbb{T}^q$ is a surface of revolution. It moves a curve around an axis of rotation. These axes are given by $1$-spheres, such that the $q$-torus is a product space $\mathbb{T}^q = \mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q$. The initial decomposition of connected commutative Lie groups can be further simplified into $M \simeq \mathbb{R}^p \times \mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q$. Recall, that the $(n-1)$-sphere is given by $\mathcal{S}^{n-1} := \{x \in \mathbb{R}^n \; \vert \; \norm{x}_2 = 1\}$. Thus, $\mathcal{S}^1$ can be embedded into $\mathbb{R}^2$. Next, we derive how many dimensions are at least needed for a suitable embedding. For each $1$-sphere we count two dimensions and for each real line accordingly one. Finally, we have to estimate from data how the decomposition may look like to yield the topology of the data manifold.

@@

\input{figure1.tex}
\subsubsection{Simplicial complexes} A \emph{simplicial complex} $K$ is a set of simplices, such that any face of a simplex of $K$ is a simplex of $K$ and the intersection of any two simplices of $K$ is either empty or a common face of both \cite[p.~11]{boissonnat2018geometric}. We require every $(k+1)$-element, with $k < n$ to be a $k$-simplex of $[\sigma] \subset K$. This defines the simplicial complex as a topological space. Examples for simplicial complexes can be found in Fig. \ref{figure2} b). These properties ensure that simplices are added to the complex in a way, such that they are \emph{glued} edge to edge, point to point, and surface to surface. Thus, every $n$-simplex has $n+1$ distinct vertices and no other $n$-simplex has the same set of vertices. This is a unique combinatorial description of vertices together with a collection of sets $\{[\sigma]_1, \cdots, [\sigma]_m\}$ of $k$-simplices, which are $k+1$ element subsets of $K$ \cite[p.~107]{hatcher2002algebraic}. This set system can be realized geometrically with an injective map into Euclidean space.

@@

The $k$-th chain groups contain two more Abelian subgroups that behave normal to it. First, the so-called cycle groups $\langle Z_k, +\rangle$, which are defined as $Z_k := \ker \partial_k = \{[\sigma] \in C_k \; \vert \; \partial_k [\sigma] = \emptyset \}$. This follows the intuition of all elements that form a loop, i.e. have the same start and end point. Some of these loops have the peculiarity of being a boundary of a subcomplex. These elements form the boundary group $\langle B_k, +\rangle$, defined by $B_k := \text{im } \partial_{k+1} = \{[\sigma] \in C_k \; \vert \; \exists [\tau] \in C_{k+1}: [\sigma] = \partial_{k+1}[\tau]\}$. The connection of these groups is illustrated in Fig. \ref{figure1} a). Now we define the homology groups as quotients of groups. Homology -- analogous to homotopy theory -- gives information about connected components, loops and higher dimensional holes in the simplicial complex:
\begin{equation}
    H_k(K) := \frac{\ker \partial_k C_k(K)}{\text{im } \partial_{k+1} C_{k+1}(K)} = \frac{Z_k(K)}{B_k(K)}.
\end{equation}

\subsubsection{Homological persistence} Examining the homology groups of a set of points gives little information about the structure of a dataset. Instead, we are interested in a parametrization of the simplicial complex as geometric realization in which the homology groups appear and disappear again. For this purpose we consider all possible subcomplexes that form a \emph{filtration} over the point set $X$. We denote $K^{\epsilon} := K^{\epsilon}(X)$. Depending on how we vary the parameter $\epsilon_i$ of the chosen simplicial complex, the following sequence is generated, connected by inclusion:
\begin{align}
    \emptyset &= K^{\epsilon_0} \subset K^{\epsilon_1} \subset \cdots \subset K^{\epsilon_{n+1}}= K,\\
    K^{\epsilon_{i+1}} &= K^{\epsilon_i} \cup [\sigma]^{\epsilon_{i+1}}, \quad \text{for} \; i \in \{0, \cdots, n-1\}.
\end{align}
An example of a filtration is given in Fig. \ref{figure1} b). The filtration has a discrete realization with a fixed $\epsilon = \epsilon_{i+1}-\epsilon_{i} = \min(x,y)$ for all $x,y \in X$. Through filtration we are able to investigate the homology groups during each step of the parameterization. We record when elements from a homology group appear and when they disappear again. Intuitively speaking, we can see when $k$-dimensional holes appear and disappear in the filtration. We call this process \emph{birth} and \emph{death} of topological features. Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}, see Fig. \ref{figure1} c)3. The Betti numbers $\beta_k$ are defined by $\text{rank } H_k$. We write $H^{\epsilon_i}_k$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\epsilon_i$. Then $H^{\epsilon_{i}}_{k} \rightarrow H^{\epsilon_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, for a proof we refer to \cite{edelsbrunner2008persistent}:
\begin{align}
0 = H^{\epsilon_{0}}_k \rightarrow H^{\epsilon_{1}}_k \rightarrow \cdots \rightarrow H^{\epsilon_{n}}_k \rightarrow H^{\epsilon_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\epsilon_i}$ or appear before and die after spanning $K^{\epsilon_{i+1}}$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\epsilon_{i},\epsilon_{j}}_k = (\beta_k^{\epsilon_{i},\epsilon_{j-1}} - \beta_k^{\epsilon_i,\epsilon_{j}})-(\beta_k^{\epsilon_{i-1},\epsilon_{j-1}}-\beta_k^{\epsilon_{i-1},\epsilon_{j}}),
\end{equation}
for the $k$-th homology group and index pairs $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \{-\infty,\infty\}$ with indices $i \leq j$. The Euclidean space is extended, as the very first connected component on the filtration remains connected. Thus, we assign to it infinite persistence, corresponding to the second coordinate $\epsilon_{j+1}$. The first term counts elements born in $K^{\epsilon_{j-1}}$ and which vanish entering $K^{\epsilon_{j}}$, while the second term counts the representatives of homology classes before $K^{\epsilon_{j}}$ and which vanish at $K^{\epsilon_{j}}$. The $k$-th persistence diagram is then defined as
\begin{equation}
\text{Ph}_{k}(X) := \left\{(\epsilon_i, \epsilon_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\epsilon_{i},\epsilon_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}

\subsubsection{Persistence landscapes} Persistence landscapes give a statistical summary of the topology of a set of points embedded in a given topological manifold \cite{bubenik2015statistical}. Looking at the points $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2}$ on the $k$-th persistence diagram $\text{Ph}_{k}(X)$, one associates a piecewise linear function $\lambda^{\epsilon_i}_{\epsilon_{j+1}}: \mathbb{R} \rightarrow [0,\infty)$ with those points:
\begin{align}
    &\text{If } x \not\in (\epsilon_i,\epsilon_{j+1}), \; \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = 0,\\
    &\text{if } x \in \left(\epsilon_{i},(\epsilon_i + \epsilon_{j+1})/2 \right], \; \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = x - \epsilon_{i} \; \text{and}\\
    &\text{if } x \in \left((\epsilon_i + \epsilon_{j+1})/2,\epsilon_{j+1} \right), \; \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = \epsilon_{j+1} - x.
\end{align}
The summaries for the general persistence diagram are the disjoint union of the $k$-th persistence diagrams $\text{Ph}(X) := \coprod_{i=0}^{k} \text{Ph}_i(X)$. A persistence landscape $\text{Pl}(X)$, contains the birth-death pairs $(\epsilon_i,\epsilon_{j+1})$, for an $i,j \in \{1, \cdots, n\}$ and is the function sequence $\Lambda_k : \mathbb{R} \rightarrow [0,\infty)$ for a $k \in \mathbb{N}$, where $\Lambda_k(x)$ denotes the $k$-th greatest value of $\lambda^{\epsilon_i}_{\epsilon_{j+1}}(x)$, see Fig. \ref{figure1} c)4. Thus, $\Lambda_k(x) = 0$ if $k > n$. $\text{Pl}(X)$ lies in a completely normed vector space, suited for statistical computations.

@@

In practice, neural networks are used with a different number of neurons per layer. However, a layer can represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map does not change. Thus, the dimension of the data manifold is equal to the width of the smallest possible layer of a neural network that is able to represent it \cite{HauserR17}. In other words, the layer with minimal width which does not loose information about the topological structure.

@@

We count the representatives of homology classes in the persistence landscapes to derive the dimension of the sought manifold. We get for $\mathcal{S}^1$:
\begin{align}
    H_0(\mathcal{S}^1) &= H_1(\mathcal{S}^1) = \mathbb{Z},\\
    H_i(\mathcal{S}^1) &= 0, \; \text{for all } i \geq 2.
\end{align}
Note, in Eq. \ref{homologykunneth} terms remain only for indices $i_{j} \in \{0,1\}$. Thus, we get
\begin{align}
    H_0(\mathbb{R}^p) &\simeq\mathbb{Z}^{p},\\
    \label{determineq}
    H_k(\mathbb{T}^{q}) &\simeq H_k(\mathcal{S}^{1}_{1} \times \cdots \times \mathcal{S}^{1}_{q}) \simeq \mathbb{Z}^{{q\choose{k}}},
\end{align}
where $p$ indicates the number of connected components. The number of local maxima for a homology group in the persistence landscape is the cardinality of the set of points $(\epsilon_i,\epsilon_{j+1})$ with derivatives $d\lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = 0 \; \text{and} \; d^2 \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) < 0$, for an example of these maximum values within a persistence landscape see Fig. \ref{figure2} b) c). Also, it is the solution for the binomial coefficient:
\begin{align}
     {q \choose k} = \frac{q}{1} \cdot \frac{q-1}{2} \cdots \frac{q-(k-1)}{k} = \prod_{i=1}^{k} \frac{q+1-i}{i}.
\end{align} 
We count in $PL(X)$ the elements of the $0$th homology group. The amount of elements from higher homology groups correspond to $q$ (Eq. \ref{determineq}) and are computed using Monte-Carlo methods for an (approximate) integer solution, see Tab. \ref{figure3}.

@@

Applying Künneth's theorem, the homology groups of the topological factor spaces could be connected with the ones of their product space. Using persistence landscapes the elements originating from some homology groups on the filtration were estimated. With numerical experiments we predicted near ideal embedding dimensions and could confirm, see Fig. \ref{figure2}, that a neural embedding above the treshold delivers a loss function on the validation data set with small fluctuations and best reconstruction results. We pose following open research questions:
\begin{itemize}
    \item The neural layers do not have the explicit structure of an Abelian Lie group. How can spherical CNNs \cite{CohenGKW18} be used to always represent a product space $\mathbb{R}^p \times \mathbb{T}^q$ from layer to layer in order to operate on the proposed Lie group?
    \item This approach can be applied to any kind of manifold, as far as its minimal representation is known. Using decompositions, one may generalize this result. What decomposition allows to neglect the connectedness? For example, the persistence diagrams could be used to identify connected components. These in return could be represented as independent connected manifolds.
    \item What do other representations of manifolds with a more complicated structure look like? For example, can manifolds like the Klein bottle be learned that are not orientable? And if so, how can one decide in advance whether or not the approximation to an orientable manifold makes more sense than to a non orientable one?
\end{itemize}
